A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-11-12 inst/manuscript/manuscript.tex
finished draft for review

  \begin{abstract}
  This is the abstract.
  \end{abstract}

@@

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as data that incorporates the
specific locations of the population units into either the design or
estimation process. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Since then, there have been several general
comparisons between design-based and model-based approaches for spatial
data (Brus and De Gruijter, 1997; Brus, 2020; Ver Hoef, 2002; Ver Hoef,
2008; Wang et al., 2012). Cooper (2006) reviews the two approaches in an
ecological context before introducing a ``model-assisted'' variance
estimator that combines aspects from each approach. In addition to
Cooper (2006), there has been substantial research and development into
estimators that use both design and model-based principles (see e.g.,
Sterba (2009), Cicchitelli and Montanari (2012), Chan-Golston et al.
(2020) for a Bayesian approach).

@@

Instead of estimating fixed but unknown parameters like a mean or total
(as in the design-based approach), the goal of model-based inference in
the spatial context is often to predict a realized variable, or value.
For example, suppose the realized mean of all population units is the
value of interest. Instead of \emph{estimating} a fixed, unknown mean,
we are \emph{predicting} the value of the mean, a random variable.
Prediction intervals are then derived using assumptions of the data
generating process. If we repeatedly generate the response values from
the same spatial process and sample, then 95\% of all 95\% prediction
intervals constructed from a procedure with appropriate coverage will
contain their respective realized means. Cressie (1993) and
Schabenberger and Gotway (2017) provide reviews of model-based
approaches for spatial data. A visual comparison of the design-based and
model-based assumptions is provided in Figure \ref{fig:fig1} (Ver Hoef
(2002) and Brus (2020) provide similar figures).

@@

We note that we only use FPBK in this paper in order to focus more on
comparing the design-based and model-based approaches. Other methods,
such as k-nearest-neighbors (Fix and Hodges, 1951; Ver Hoef and
Temesgen, 2013), random forest (Breiman, 2001), Bayesian models
(Chan-Golston et al., 2020), among others, could also be used to obtain
predictions for a mean or total from spatially correlated responses of a
finite population. We choose to use FPBK because it is faster than a
Bayesian approach and it was developed with theoretically-based variance
estimators of means and totals for spatial data, whereas random forests
and k-nearest-neighbors use ad-hoc variance estimators in most cases
(Ver Hoef and Temesgen, 2013); additionally, FBPK outperformed the other
methods in most scenarios.

@@

The Environmental Protection Agency (EPA), states, and tribes
periodically conduct National Aquatic Research Surveys (NARS) in the
United States to assess the water quality of various bodies of water. We
will use the 2012 National Lakes Assessment (NLA), which measures
various aspects of lake health and quality in lakes in the contiguous
United States, to obtain an interval for mean mercury concentration.
Although we know the true mean mercury concentration values for the 986
lakes from the 2012 NLA, we will explore whether or not we obtain an
adequately precise estimate for the realized mean mercury concentration
if we sample only 100 of the 986 lakes.

\begin{figure}
\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/figdata-1} \caption{Population distribution of mercury concentration for 986 lakes in the contiguous United States.}\label{fig:figdata}
\end{figure}

@@

@@

Table \ref{tab:appliedtab} shows the application of a design-based
analysis of an IRS sample, a model-based analysis of an IRS sample, a
design-based analysis of a GRTS sample, and a model-based analysis of a
GRTS sample. For all four analyses, the true realized mean mercury
concentration is within the bounds of the 95\% intervals. However, we
should not generalize the results of this particular realization to any
other data set or even to other potential samples of this data set.

But, we do note a couple of patterns. The design-based IRS analysis
shows the largest standard error: a likely reason is that this is the
only approach that does not incorporate any spatial information
regarding mercury concentration across the contiguous United States. We
also see that both approaches using the GRTS sample have a lower
standard error than the both approaches using the IRS sample. We would
expect this to be the case for most samples because mercury
concentration exhibits spatial patterning, so a spatially balanced
sample should usually yield a lower standard error.

@@

The design-based and model-based approaches to inference are
fundamentally different paradigms by which to samples are selected and
data are analyzed. The design-based approach incorporates randomness
through sampling to estimate a population parameter. The model-based
approach incorporates randomness through distributional assumptions to
predict the realized value of a random variable. Though these approaches
have often been compared in the literature both from theoretical and
analytical perspectives, our contribution lies in studying them in a
spatial context while implementing spatially balanced sampling. Aside
from the theoretical differences described, a few analytical findings
are particularly notable: the design decision (GRTS vs IRS) seems much
more important than the analysis decision (design-based vs model-based);
independent of the analysis approach, there is no reason to prefer IRS
over GRTS for spatial data -- GRTS tends to perform at least as well as
IRS when there is no spatial correlation and increasingly better than
IRS as the strength of spatial correlation increases; the gap in
relative efficiency between GRTS-design and GRTS-model widens as the
strength of spatial correlation increases; and when the data are skewed,
interval coverage for all approaches improves both as the sample size
increases and as the strength of correlation increases.

There are several benefits and drawbacks of the design-based and
model-based approaches for spatial data, some of which we have not yet
discussed but are worthy of consideration in future research.
Design-based approaches are often computationally efficient, while
model-based estimation of covariance parameters can be computationally
burdensome, especially for likelihood-based methods such as REML that
rely on inverting a covariance matrix. The design-based approach also
more naturally handles binary data, free from the more complicated
logistic regression formulation commonly used to handle binary data in a
model-based approach. The model-based approach, however, can more
naturally quantify the relationship between covariates (predictor
variables) and the response variable. The model-based approach also
yields estimated spatial covariance parameters, which help better
understand the process of study. Model selection is also possible using
model-based approaches and criteria such as likelihood ratio tests or
AIC (Akaike, 1974). Model-based approaches are capable of more efficient
small-area estimation than design-based approaches by leveraging
distributional assumptions in areas with few observed sites. Model-based
approaches can also compute site-by-site predictions at unobserved
locations and use them to construct informative visualizations. The
benefits and drawbacks of both approaches, alongside our theoretical and
analytical comparisons, should be heavily considered when choosing among
them. This is especially true from an analysis perspective, as we found
that using a spatially balanced sampling algorithm benefits both
design-based and model-based analyses.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-akaike1974new}{}%
Akaike, H., 1974. A new look at the statistical model identification.
IEEE transactions on automatic control 19, 716--723.

@@

\leavevmode\hypertarget{ref-breiman2001random}{}%
Breiman, L., 2001. Random forests. Machine learning 45, 5--32.

\leavevmode\hypertarget{ref-brus1997random}{}%
Brus, D., De Gruijter, J., 1997. Random sampling or geostatistical
modelling? Choosing between design-based and model-based sampling
strategies for soil (with discussion). Geoderma 80, 1--44.

\leavevmode\hypertarget{ref-brus2020statistical}{}%
Brus, D.J., 2020. Statistical approaches for spatial sample survey:
Persistent misconceptions and new developments. European Journal of Soil
Science.

@@

\leavevmode\hypertarget{ref-cooper2006sampling}{}%
Cooper, C., 2006. Sampling and variance estimation on continuous
domains. Environmetrics: The official journal of the International
Environmetrics Society 17, 539--553.

@@

\leavevmode\hypertarget{ref-de1990model}{}%
De Gruijter, J., Ter Braak, C., 1990. Model-free estimation from spatial
samples: A reappraisal of classical sampling theory. Mathematical
geology 22, 407--415.

@@

\leavevmode\hypertarget{ref-fix1951discriminatory}{}%
Fix, E., Hodges, J.L., 1951. Discriminatory analysis, nonparametric
discrimination: Consistency properties. USAF School of Aviation
Medicine.

@@

\leavevmode\hypertarget{ref-harville1977maximum}{}%
Harville, D.A., 1977. Maximum likelihood approaches to variance
component estimation and to related problems. Journal of the American
statistical association 72, 320--338.

@@

\leavevmode\hypertarget{ref-horvitz1952generalization}{}%
Horvitz, D.G., Thompson, D.J., 1952. A generalization of sampling
without replacement from a finite universe. Journal of the American
statistical Association 47, 663--685.

@@

\leavevmode\hypertarget{ref-sterba2009alternative}{}%
Sterba, S.K., 2009. Alternative model-based and design-based frameworks
for inference from samples to populations: From polarization to
integration. Multivariate behavioral research 44, 711--740.

@@

\leavevmode\hypertarget{ref-stevens2004spatially}{}%
Stevens, D.L., Olsen, A.R., 2004. Spatially balanced sampling of natural
resources. Journal of the american Statistical association 99, 262--278.

@@

\leavevmode\hypertarget{ref-ver2013comparison}{}%
Ver Hoef, J.M., Temesgen, H., 2013. A comparison of the spatial linear
model to nearest neighbor (k-NN) methods for forestry applications. PloS
one 8, e59129.

\leavevmode\hypertarget{ref-wang2013design}{}%
Wang, J.-F., Jiang, C.-S., Hu, M.-G., Cao, Z.-D., Guo, Y.-S., Li, L.-F.,
Liu, T.-J., Meng, B., 2013. Design-based spatial sampling: Theory and
implementation. Environmental modelling \& software 40, 280--288.

>>>>

  \begin{abstract}
  The design-based and model-based approaches to frequentist statistical
  inference rest on fundamentally different foundations. In the
  design-based approach, inference depends on random sampling. In the
  model-based approach, inference depends on distributional assumptions.
  In this manuscript, we compare the approaches for finite population
  spatial data. We first provide relevant background for design-based
  and model-based approaches to finite population patial data. Then we
  use a simulation study and an analysis of real mercurcy concentration
  data to compare them numerically. We find that sampling plans that
  incorporate spatial locations (spatially balanced samples) perform
  better than sampling plans ignoring spatial locations (non-spatially
  balanced samples), regardless of whether design-based or model-based
  approaches were used to analyze the data. We also find that within
  sampling plans, model-based approaches tend to outperform design-based
  approaches, even for skewed data. This gap in performance is small
  when spatially samples are used but large when non-spatially balanced
  samples are used.
  \end{abstract}

@@

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as data that incorporates the
specific locations of the population units into either the design or
estimation process. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Since then, there have been several general
comparisons between design-based and model-based approaches for spatial
data (Brus and De Gruijter, 1997; Brus, 2021; Ver Hoef, 2002; Ver Hoef,
2008; Wang et al., 2012). Cooper (2006) reviews the two approaches in an
ecological context before introducing a ``model-assisted'' variance
estimator that combines aspects from each approach. In addition to
Cooper (2006), there has been substantial research and development into
estimators that use both design and model-based principles (see e.g.,
Sterba (2009), Cicchitelli and Montanari (2012), Chan-Golston et al.
(2020) for a Bayesian approach).

@@

Instead of estimating fixed but unknown parameters like a mean or total
(as in the design-based approach), the goal of model-based inference in
the spatial context is often to predict a realized variable, or value.
For example, suppose the realized mean of all population units is the
value of interest. Instead of \emph{estimating} a fixed, unknown mean,
we are \emph{predicting} the value of the mean, a random variable.
Prediction intervals are then derived using assumptions of the data
generating process. If we repeatedly generate the response values from
the same spatial process and sample, then 95\% of all 95\% prediction
intervals constructed from a procedure with appropriate coverage will
contain their respective realized means. Cressie (1993) and
Schabenberger and Gotway (2017) provide reviews of model-based
approaches for spatial data. A visual comparison of the design-based and
model-based assumptions is provided in Figure \ref{fig:fig1} (Ver Hoef
(2002) and Brus (2021) provide similar figures).

@@

We note that we only use FPBK in this paper in order to focus more on
comparing the design-based and model-based approaches. Other methods,
such as k-nearest-neighbors (Fix and Hodges, 1989; Ver Hoef and
Temesgen, 2013), random forest (Breiman, 2001), Bayesian models
(Chan-Golston et al., 2020), among others, could also be used to obtain
predictions for a mean or total from spatially correlated responses of a
finite population. We choose to use FPBK because it is faster than a
Bayesian approach and it was developed with theoretically-based variance
estimators of means and totals for spatial data, whereas random forests
and k-nearest-neighbors use ad-hoc variance estimators in most cases
(Ver Hoef and Temesgen, 2013); additionally, FBPK outperformed the other
methods in most scenarios.

@@

The Environmental Protection Agency (EPA), states, and tribes
periodically conduct National Aquatic Research Surveys (NARS) in the
United States to assess the water quality of various bodies of water. We
will use the 2012 National Lakes Assessment (NLA), which measures
various aspects of lake health and quality in lakes in the contiguous
United States, to study mercury concentration. Although we know the true
mean mercury concentration values for the 986 lakes from the 2012 NLA,
we will explore whether or not we obtain an adequately precise estimate
for the realized mean mercury concentration if we sample only 100 of the
986 lakes.

@@

\begin{figure}

{\centering \includegraphics[width=0.49\linewidth]{manuscript_files/figure-latex/figdata-1} \includegraphics[width=0.49\linewidth]{manuscript_files/figure-latex/figdata-2}

}

\caption{Population distribution of mercury concentration (hg) for 986 lakes in the contiguous United States in a spatial layout (left) and a histogram (right).}\label{fig:figdata}
\end{figure}

@@

We selected a single IRS sample and a single GRTS sample and estimated
the mean mercury concentration and its standard error using using
design-based and model-based approaches; Table \ref{tab:appliedtab}
shows the results. For all four sampling-analysis combinations, the true
realized mean mercury concentration is within the bounds of the 95\%
intervals. However, we should not generalize these results to any other
data or even to other samples from these data. But, we do note a couple
of patterns. The design-based IRS analysis shows the largest standard
error: a likely reason is that this is the only approach that does not
incorporate any spatial information regarding mercury concentration
across the contiguous United States. We also see that both approaches
using the GRTS sample have a lower standard error than the both
approaches using the IRS sample. We would expect this to be the case for
most samples because mercury concentration exhibits spatial patterning,
so a spatially balanced sample should usually yield a lower standard
error.

To better understand the dependence structure in mercury concentration,
the empirical semivariogram and corresponding fit of the model-based
approaches can be visualized. The empirical semivariogram quantifies the
halved squared differences (semivariance) among response values at
different distances apart. If a process exhibits strong spatial
dependence, the empirical semivariogram will have small values at small
distances and large values at large distances. Figure \ref{fig:figsv}
shows the empirical semivariogram for GRTS-Model, displaying the average
semivariance for several distances. Overlain onto Figure \ref{fig:figsv}
is the estimated semivariance obtained using the covariance parmaeters
from the REML fit of GRTS-Model. Figure \ref{fig:figsv} provides
evidence that there is strong correlation in mercury concentration among
the sites.

\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{manuscript_files/figure-latex/figsv-1}

}

\caption{The empirical semivariogram (black circles) of mercury concentration against the REML fit using the estimated covariance parameters (black line) from GRTS-Model.}\label{fig:figsv}
\end{figure}

@@

The design-based and model-based approaches to inference are
fundamentally different paradigms by which samples are selected and data
are analyzed. The design-based approach incorporates randomness through
sampling to estimate a population parameter. The model-based approach
incorporates randomness through distributional assumptions to predict
the realized values of a random process. Though these approaches have
often been compared in the literature both from theoretical and
analytical perspectives, our contribution lies in studying them in a
spatial context while implementing spatially balanced sampling. Aside
from the theoretical differences described, a few analytical findings
from the simulation study are particularly notable. First, the sampling
decision (GRTS vs IRS) is most important when using a design-based
analysis. Though GRTS-Model still outperformed IRS-Model, the
model-based analysis mitigated much of the inefficiency of the IRS
sample. Second, independent of the analysis approach, there is no reason
to use IRS over GRTS for sampling spatial data, as GRTS-Design and
GRTS-Model generally performed at least as well as their IRS
counterparts when there was no spatial correlation and noticeably better
than there IRS counterparts when there was spatial correlation. Third,
The stronger the spatial correlation, the larger the gap in rMS(P)E
between IRS-Design and the other sampling-analysis combinations. Fourth
and finally, interval coverage for the normal response was very close to
95\% for all sample sizes, while interval coverage for the lognormal
response was not very close to 95\% until \(n = 200\).

There are several benefits and drawbacks of the design-based and
model-based approaches for spatial data, some of which we have not yet
discussed but are worthy of consideration in future research.
Design-based approaches are often computationally efficient, while
model-based estimation of covariance parameters can be computationally
burdensome, especially for likelihood-based methods such as REML that
rely on inverting a covariance matrix. The design-based approach also
more naturally handles binary data, free from the more complicated
logistic regression formulation commonly used to handle binary data in a
model-based approach. The model-based approach, however, can more
naturally quantify the relationship between covariates (predictor
variables) and the response variable. The model-based approach also
yields estimated spatial covariance parameters, which help better
understand the process of study. Model selection is also possible using
model-based approaches and criteria such as cross validation, likelihood
ratio tests, or AIC (Akaike, 1974). Model-based approaches are capable
of more efficient small-area estimation than design-based approaches by
leveraging distributional assumptions in areas with few observed sites.
Model-based approaches can also compute site-by-site predictions at
unobserved locations and use them to construct informative
visualizations. The benefits and drawbacks of both approaches, alongside
our theoretical and analytical comparisons, should be seriously
considered when choosing among them. This is especially true from an
analysis perspective, as we found that using a spatially balanced
sampling algorithm benefits both design-based and model-based analyses.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-akaike1974new}{}%
Akaike, H., 1974. A new look at the statistical model identification.
IEEE Transactions on Automatic Control 19, 716--723.

@@

\leavevmode\hypertarget{ref-breiman2001random}{}%
Breiman, L., 2001. Random forests. Machine Learning 45, 5--32.

\leavevmode\hypertarget{ref-brus1997random}{}%
Brus, D., De Gruijter, J., 1997. Random sampling or geostatistical
modelling? Choosing between design-based and model-dased sampling
strategies for soil (with discussion). Geoderma 80, 1--44.

\leavevmode\hypertarget{ref-brus2021statistical}{}%
Brus, D.J., 2021. Statistical approaches for spatial sample survey:
Persistent misconceptions and new developments. European Journal of Soil
Science 72, 686--703.

@@

\leavevmode\hypertarget{ref-cooper2006sampling}{}%
Cooper, C., 2006. Sampling and variance estimation on continuous
domains. Environmetrics 17, 539--553.

@@

\leavevmode\hypertarget{ref-de1990model}{}%
De Gruijter, J., Ter Braak, C., 1990. Model-free estimation from spatial
samples: A reappraisal of classical sampling theory. Mathematical
Geology 22, 407--415.

@@

\leavevmode\hypertarget{ref-fix1989discriminatory}{}%
Fix, E., Hodges, J.L., 1989. Discriminatory analysis. Nonparametric
discrimination: Consistency properties. International Statistical
Review/Revue Internationale de Statistique 57, 238--247.

@@

\leavevmode\hypertarget{ref-harville1977maximum}{}%
Harville, D.A., 1977. Maximum likelihood approaches to variance
component estimation and to related problems. Journal of the American
Statistical Association 72, 320--338.

@@

\leavevmode\hypertarget{ref-horvitz1952generalization}{}%
Horvitz, D.G., Thompson, D.J., 1952. A generalization of sampling
without replacement from a finite universe. Journal of the American
Statistical Association 47, 663--685.

@@

\leavevmode\hypertarget{ref-sterba2009alternative}{}%
Sterba, S.K., 2009. Alternative model-based and design-based frameworks
for inference from samples to populations: From polarization to
integration. Multivariate Behavioral Research 44, 711--740.

@@

\leavevmode\hypertarget{ref-stevens2004spatially}{}%
Stevens, D.L., Olsen, A.R., 2004. Spatially balanced sampling of natural
resources. Journal of the American Statistical Association 99, 262--278.

@@

\leavevmode\hypertarget{ref-ver2013comparison}{}%
Ver Hoef, J.M., Temesgen, H., 2013. A comparison of the spatial linear
model to nearest neighbor (k-NN) methods for forestry applications. PlOS
ONE 8, e59129.

\leavevmode\hypertarget{ref-wang2013design}{}%
Wang, J.-F., Jiang, C.-S., Hu, M.-G., Cao, Z.-D., Guo, Y.-S., Li, L.-F.,
Liu, T.-J., Meng, B., 2013. Design-based spatial sampling: Theory and
implementation. Environmental Modelling \& Software 40, 280--288.
